Shallow EDSLs and Object-Oriented Programming

### 2016-06-30 paper/sections/Introduction.tex
paper
intro

\noindent Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in Object-Oriented Programming (OOP). Would this OO approach be
classified as a shallow or deep embedding? We believe arguments can be
made both ways. 
%The {\sc Composite} or {\sc Interpreter}
%patterns are normally accepted to provide a way to encode ASTs. Thus, 
%one possible interpretation is that \emph{according to definition of deep embedding
%  above, the OO approach classifies as a deep
%  embedding}.

@@

In this paper, to avoid ambiguity, we prefer a more precise meaning
for shallow embeddings as EDSLs implemented using \emph{procedural abstraction}~\cite{reynolds94proceduralabstraction}. Such
interpretation arises from the domain of shallow EDSLs being
typically a function, and procedural abstraction being a way to encode
data abstractions using functions. As Cook~\cite{cook09abstraction} argued,
procedural abstraction is also the essence of OOP.
Thus, according to our definition, the implementation of a shallow
EDSL in OOP languages should simply correspond to a standard
object-oriented program.

@@

The main goal of this paper is to argue that OOP languages have
advantages for the implementation of shallow embeddings. 
%This should
%not came as a surprise, as OOP languages have evolved over more than 50
%years to improve the use of procedural abstraction. 
In order to understand how OOP can help with shallow embeddings, it is
helpful to
first review some of the limitations commonly
found in the literature:
\begin{enumerate}

@@

\noindent Both limitations are often used as motivators to switch to deep embeddings.

@@

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala\footnote{Available online: \url{https://github.com/wxzh/shallow-dsl}}.
From the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages over the Haskell version.

>>>>

\noindent Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in Object-Oriented Programming (OOP). Would this OO approach be
classified as a shallow or deep embedding? Some authors working on
OOP EDSLs ~\cite{} consider a {\sc Composite} to be a deep
embedding. Some other authors~\cite{gibbons2014folding} 
consider implementations using tuples and/or the  {\sc Composite}
pattern to be a shallow embedding.

%The {\sc Composite} or {\sc Interpreter}
%patterns are normally accepted to provide a way to encode ASTs. Thus, 
%one possible interpretation is that \emph{according to definition of deep embedding
%  above, the OO approach classifies as a deep
%  embedding}.

@@

In this paper, we avoid ambiguitity by using a more precise meaning
for shallow embeddings as EDSLs implemented using \emph{procedural abstraction}~\cite{reynolds94proceduralabstraction}. Such
interpretation arises naturally from the domain of shallow EDSLs being
typically functions, and procedural abstraction being a way to encode
data abstractions using functions. As Cook~\cite{cook09abstraction} argued,
procedural abstraction is also the essence of OOP.
Thus, according to our definition, the implementation of a shallow
EDSL in OOP languages should simply correspond to a standard
object-oriented program.

@@

The main goal of our paper is to show that OOP languages have
advantages for the implementation of shallow embeddings. 
It is perhaps due to those advantages that some authors 
have considered  {\sc Composite}-based implementations to be deep
embeddings, rather than shallow. However, we argue that the OOP
mechanisms do not change the essence of the implementation, which is
still shallow (i.e. using procedural abstraction). To understand how
OOP is helpful for shallow embeddings, we should first look at the
limitations of shallow embeddings commonly found in the literature:
\begin{enumerate}

@@

@@

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala\footnote{Available online: \url{https://github.com/wxzh/shallow-dsl}}.
From the \emph{modularity} point of view the
Scala version has clear advantages over the Haskell version.
